---
title: "Sweet Spot Analysis"
author: "Alice Tivarovsky"
date: "2025-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load Libraries and Themes

```{r libraries}
pacman::p_load(
  tidyverse,
  dbplyr,
  DBI,
  odbc, 
  lubridate,
  plotly
  
)

theme_set(theme_bw())
set.seed(1234)

options(scipen=999) # eliminates scientific notation formatting

```


## To Do

- pull in ACG
- pull in cost data from claims
- pull in Epic
- add logic to code new/existing patient


## Questions

- What, on average, is the difference in cost between new and established patients?


## Import and Tidy Data

### Connect to Servers/Databases

```{r odbc connect}
mcanalytics <- DBI::dbConnect(odbc::odbc()
                        ,Driver = "ODBC Driver 17 for SQL Server"
                        ,Server = "DB-PROD-MSO"
                        ,Database = "MCANALYTICS"
                        ,Trusted_Connection = "Yes")

mcis <- DBI::dbConnect(odbc::odbc()
                        ,Driver = "ODBC Driver 17 for SQL Server"
                        ,Server = "SQL12"
                        ,Database = "MCIS"
                        ,Trusted_Connection = "Yes")

altamed_dwh <- DBI::dbConnect(odbc::odbc()
                        ,Driver = "ODBC Driver 17 for SQL Server"
                        ,Server = "DB-PROD-DWH"
                        ,Database = "ALTAMED_DWH"
                        ,Trusted_Connection = "Yes")

acg <- DBI::dbConnect(odbc::odbc()
                        ,Driver = "ODBC Driver 17 for SQL Server"
                        ,Server = "DB-PROD-DWH"
                        ,Database = "ACG"
                        ,Trusted_Connection = "Yes")

wakely <- DBI::dbConnect(odbc::odbc()
                        ,Driver = "ODBC Driver 17 for SQL Server"
                        ,Server = "DB-PROD-DWH"
                        ,Database = "Wakely"
                        ,Trusted_Connection = "Yes")

datalake <- DBI::dbConnect(odbc::odbc()
                        ,Driver = "ODBC Driver 17 for SQL Server"
                        ,Server = "DB-PROD-REPORT"
                        ,Database = "Datalake"
                        ,Trusted_Connection = "Yes")

clarity <- DBI::dbConnect(odbc::odbc()
                        ,Driver = "ODBC Driver 17 for SQL Server"
                        ,Server = "DB-PROD-CLARITY"
                        ,Database = "CLARITY"
                        ,Trusted_Connection = "Yes")


```


### Import Tables

Define interval start date and end date. 
```{r}
start_date <- '2024-03-01'
end_date <- '2025-03-01'

```

#### Epic 

Epic table to crosswalk Patient IDs to EZ Cap IDs
```{r}
epic_patient_for_app_crosswalk <-
  mcis |>
  tbl(in_schema("dbo", "Epic_Patient_for_App")) |>
  select(c(PatientSID, EZ_MembID, PatientEpicId, PrimaryMrn)) |>
#  filter(PatientSID %in% sample_patient_ids) |>
  collect() |>
  janitor::clean_names()

```


#### MCANALYTICS Tables 

`MCANALYTICS` claims tables
```{r}
claim_header <- 
  mcanalytics |> 
  tbl(in_schema("data", "tblClaim_Header")) |> 
  select(c(ClaimNo, DateCreated, DateFrom, MemberId, CompanyId, TotalNet, PlaceofSvcCode, StatusCode)) |> 
  filter(DateFrom >= start_date & DateFrom <= end_date) |> 
  filter(CompanyId %in% c("ALTACHLA", "ALTAMED", "ALTANET", "ALTAOC") & StatusCode == '9') |> 
  collect() |> 
  janitor::clean_names() |> 
  filter(!is.na(claim_no))
  		
claim_details <-
  mcanalytics |>
  tbl(in_schema("data", "tblClaim_Details")) |>
  select(c(ClaimNo, CompanyId, DateFrom, NetPaid
           , ProcCode)) |>
  filter(DateFrom >= start_date & DateFrom <= end_date) |>
  filter(CompanyId %in% c("ALTACHLA", "ALTAMED", "ALTANET", "ALTAOC")) |>
  collect() |>
  janitor::clean_names() |>
  select(-date_from) |>
  filter(!is.na(claim_no))

```


`MCANALYTICS` `MMS01` table 
```{r}
tbl_mms01 <- 
  mcanalytics |> 
  tbl(in_schema("rptstg", "tblMMS01")) |> 
  select(c(CompanyID, YOS, QOS, MOS, MOS_Date, LobCode, LobName, HealthPlanContractCode, HealthPlanContractDescription, HealthPlanName, Memberid, MemberName, MemberZipCode
        ,BenefitOption
        ,MedicalProgram
        ,Sex
        ,AgeGrp
        ,Age
        ,AidCat
        ,AidCode
        ,MemberCounty
        ,MemberRegion
        ,MMS
        ,Autono
        ,MPINO
        ,DataThrough
        ,MemberRegionDescription
        ,Altamed
        ,AgeGroup2
        ,Program
        ,Clinic
        ,RiskGroup
        ,PCPCounty
        ,PCPGeographicRegion
        ,Region
        ,MMS_NEW
        ,MMS_TERM
        ,MMS_Exist
        ,MMS_Exist_90day
)) |> 
  collect() |> 
  janitor::clean_names() |> 
  filter(yos %in% c(2023, 2024, 2025))

```


DimPatient Table to crosswalk patient IDs
```{r}
dim_patient <- 
  altamed_dwh |> 
  tbl(in_schema("dbo", "DimAllPatient")) |> 
  select(CurrentPatientSID, AllPatientID) |> 
  collect() |> 
  janitor::clean_names()

memb_company <- 
  datalake |> 
  tbl(in_schema("EZCap", "MEMB_COMPANY")) |> 
  select(c(MEMB_KEYID, MEMB_MPI_NO, MEMBID, COMPANY_ID)) |> 
  collect() |> 
  janitor::clean_names()

patient_id_crosswalk <- 
  dim_patient |> 
  left_join(memb_company, by = c("all_patient_id" = "memb_mpi_no")) |> 
  filter(!is.na(membid)) |> 
  select(c(current_patient_sid, membid, company_id)) |> 
  unique()

```



## Data Exploration
```{r}
tbl_mms01 |> 
  group_by(mms_new) |> 
  summarise(n = n())

```

Patient 90423751G, flagged as a new member 2024-02 (AltaOC) and  2024-11 (AltaNet), started as AltaOC then has first AltaNet observation, then every Altanet observation is repeated with AltaOC
```{r}
tbl_mms01 |> 
  filter(mms_new == 1) |> 
  filter(memberid == '90423751G')

tbl_mms01 |> 
  filter(memberid == '90423751G') |> 
  arrange(mos_date) |> 
  select(c(memberid, company_id, region, mos, mms_new, mms_exist, mms_exist_90day))

```


```{r}
# looks like mms_exist_90day might be the correct variable
mms_90_day_yes <- 
  tbl_mms01 |> 
  filter(mms_exist_90day == 0) |> 
  arrange(memberid)

# member John Le 00000249628
tbl_mms01 |> 
  filter(memberid == '00000249628') # new member Jun 2023

# member antonio hernandez, new 2024-04
tbl_mms01 |> 
  filter(memberid == '00000239639') |> 
  select(c(memberid, company_id, region, mos, mms_new, mms_exist, mms_exist_90day))

# member Ana Lopez Melendez, new AltaMed Feb 2024
tbl_mms01 |> 
  filter(memberid == '00000170866') |> 
  select(c(memberid, company_id, region, mos, mms_new, mms_exist, mms_exist_90day))

# 00000178238	MONARREZ, GERMAN, Jan 2025
tbl_mms01 |> 
  filter(memberid == '00000178238') |> 
  select(c(memberid, company_id, region, mos, mms_new, mms_exist, mms_exist_90day)) # checks out, mms_90_day = 0 when mms_new = 1

```

Find new vs established members for March 2024. 
```{r}
tbl_mms01_mar_2024 <- 
  tbl_mms01 |> 
  filter(mos == 202403 & company_id %in% c('ALTAMED', 'ALTAOC', 'ALTACHLA'))

```

Add Patient IDs, then pull in ACG, then pull in claims. 
```{r}
# using Epic tables doesn't work as well 
# tbl_mms01_mar_2024 |> 
#   left_join(epic_patient_for_app_crosswalk |> select(c(patient_sid, ez_memb_id)), by = c("memberid" = "ez_memb_id")) |> 
#   filter(!is.na(patient_sid)) # this loses 100k patients that don't have a patient SID match

tbl_mms01_mar_2024 <- 
  tbl_mms01_mar_2024 |> 
  left_join(patient_id_crosswalk |> filter(company_id != "ALTANET") |> select(-company_id), by = c("memberid" = "membid")) |> 
  filter(!is.na(current_patient_sid))

```

Claims at the patient level, start March 2024, end March 2025
```{r}
claim_header_by_patient <- 
  claim_header |> 
  filter(date(date_from) >= start_date & date(date_from) <= end_date) |> 
  mutate(month_year_from = floor_date(as.Date(date_from), "month")) |> 
  group_by(member_id, month_year_from) |> 
  summarise(total_cost = sum(total_net))
  
```

Plot averages
```{r}
plot_cost_by_month <- 
  claim_header_by_patient |> 
  left_join(tbl_mms01_mar_2024 |> select(c(memberid, mms_exist_90day)), by = c("member_id" = "memberid")) |> 
  filter(!is.na(mms_exist_90day)) |> 
  mutate(mms_exist_90day = as.factor(mms_exist_90day)) |> 
  group_by(mms_exist_90day, month_year_from) |> 
  summarise(average_cost = mean(total_cost)) 

plot_cost_by_month |> 
  ggplot(aes(x = month_year_from, y = average_cost, fill = mms_exist_90day)) + 
  geom_point(aes(color = mms_exist_90day)) + 
  geom_smooth(aes(color = mms_exist_90day), se = FALSE)

```


